A simulation model of DNA methylation data

### 2015-03-25 Rmarkdown/methsim.Rmd
Describe extrapolating LOR distributions.

[^WGBSSuite]: Unfortunately, `WGBSSuite` is not available as an R package, which is the "fundamental unit of reproducible R code" ([http://r-pkgs.had.co.nz/](http://r-pkgs.had.co.nz/)) and would greatly simplify the installation and use of the software. Furthermore, no licence file is included in the download, meaning that it is unclear how the user is permitted to use `WGBSSuite` and whether they may modify or redistribute the code. (__TODO: Email the developer, Owen Rackham, asking him to add a licence. He is considering my request.__)

@@

I do not think this a useful or necessary step. In fact, I think it is counterproductive. While in truth the set of methylation loci do vary between samples due to genetic variation, it is a reasonable approximation to consider the positions of these loci as fixed. If sequence variation is required then it is easily accommodations by sampling from the set of methylation loci in the reference genome. And it is counterproductive because the aim of the simulation model is to realistically simulate methylation _levels_, not the _locations_ of these loci. Furthermore, almost all downstream analyses are _reference-based_ (__TODO: Cite relevant section__), so it is desirable to know how these methods perform with respect to the relevant reference genome, not some simulated genome in which the location of methylation loci vary from simulation to simulation.

@@

For simplicity, `methsim` assumes that all $n_{\tilde{h}}$ realisations have the same parameters (described below; __TODO: Discuss implications of this assumption somewhere below__).

@@

1. Identify partially methylated regions. A summary statistic, $\alpha$, (__TODO: What is $\alpha$__) which is based on the $\beta$-values in a sliding window of 100 CpGs, is used to identify PMRs. Briefly, a two-state hidden Markov model is fit to the $\alpha$ values to identify PMRs and non-PMRs.
2. Identify UMRs and LMRs. The PMRs are masked from the genome and simple heuristics are used to identify UMRs and LMRs based on the average $\beta$-values in a window and the number of CpGs in the window.

@@

The second step requires further explanation. If there are an insufficient number of observations with the given $IPD$-region combination, then, by default, `methsim` samples $\Psi_{(i, i + 1)}$ from a Gaussian distribution with mean equal to __TODO: What?__ and standard deviation equal to __TODO: What?__. __TODO: Explain how these means and standard deviations are estimated using loess.__ By default, if there are less than __TOOD: how many?__ $LOR$ with the given $IPD(i, i + 1), r_{i}$ combination then the Normal-approximation is used. The user may specify a different method for dealing with these rarer $IPD$-region combinations.

@@

With all the above limitations in mind, `methsim` still uses the simple strategy of sampling the $\mathbf{W}$ from $w_{(i, \ldots, i + m - 1) | r_{(i, \ldots, i + m - 1)}}$. At present I do not have a better alternative and, as we shall see, the current procedure still procedure still produced realistic simulated data.

>>>>

[^WGBSSuite]: Unfortunately, `WGBSSuite` is not available as an R package, which is the "fundamental unit of reproducible R code" ([http://r-pkgs.had.co.nz/](http://r-pkgs.had.co.nz/)) and would greatly simplify the installation and use of the software. Furthermore, no licence file is included in the download, meaning that it is unclear how the user is permitted to use `WGBSSuite` and whether they may modify or redistribute the code. (__TODO: Update if I received a response. I emailed the developer, Owen Rackham, asking him to add a licence. He is considering my request.__)

@@

I do not think this a useful or necessary step. In fact, I think it is counterproductive. While in truth the set of methylation loci do vary between samples due to genetic variation, it is a reasonable approximation to consider the positions of these loci as fixed. If sequence variation is required then it is easily accommodations by sampling from the set of methylation loci in the reference genome. And it is counterproductive because the aim of the simulation model is to realistically simulate methylation _levels_, not the _locations_ of these loci. Furthermore, almost all downstream analyses are _reference-based_ (see Section \ref{sec:methylated_or_unmethylated}), so it is desirable to know how these methods perform with respect to the relevant reference genome, not some simulated genome in which the location of methylation loci vary from simulation to simulation.

@@

For simplicity, `methsim` assumes that all $n_{\tilde{h}}$ realisations have the same parameters (discussed below; __TODO: Discuss implications of this assumption somewhere below__).

@@

1. Identify partially methylated regions. A summary statistic, $\alpha$, which is based on the $\beta$-values in a sliding window of 100 CpGs, is used to identify PMRs. Briefly, a two-state hidden Markov model is fit to the $\alpha$ values to identify PMRs and non-PMRs.
2. Identify UMRs and LMRs. The PMRs are masked from the genome and simple heuristics are used to identify UMRs and LMRs based on the average $\beta$-values in a window and the number of CpGs in the window.

@@

The second step requires further explanation. If there are an insufficient number of observations with the given $IPD$-region combination, then, by default, `methsim` samples $\Psi_{(i, i + 1)}$ from a Gaussian distribution[^loess_params]. By default, if there are less than __TOOD: how many?__ $LOR$ with the given combination of region-type and IPD then the Gaussian-approximation is used. The user may specify a different method for dealing with these rarer $IPD$-region combinations and, indeed, different methods are explored in the __RESULTS SECTION__.

[^loess_params]: The mean, $\mu$, and standard deviation, $\sigma$, of this Gaussian distribution depend on both $IPD$ and $r_{i}$. A loess fit of $mean(LOR)$ (resp. $sd(LOR)$) as a function of $IPD$ is performed, stratified by region type. The values of $\mu$ (resp. $\sigma$) are then obtained by extrapolating the loess fit (__TODO: Update if changes made to `methsim` defaults__).

@@

With all the above limitations in mind, `methsim` still uses the simple strategy of sampling the $\mathbf{W}$ from $w_{(i, \ldots, i + m - 1) | r_{(i, \ldots, i + m - 1)}}$. At present I do not have a better alternative and, as we shall see, the current procedure still procedure still produced realistic simulated data. In the __RESULTS SECTION__ I also explore a simulation that simulates $n_{\bar{h}}$ independent realisations of $\mathbf{Z}$ where $n_{\tilde{h}} < n_{\bar{h}} \\l n_{h}$ and uniformly sets $\mathbf{W_{r_{i}, \cdot}} = (\frac{1}{n_{\bar{h}}, \ldots, \frac{1}{n_{\bar{h}})$.
